\SW: II. New Gravitational Lens Candidates from the CFHTLS Discovered through Citizen Science

### 2015-05-09 doc/sw-cfhtls.tex

We extracted contiguous cutouts of size 82\arcsec\ (440 pixels), including
overlapping region of 10\arcsec (54 pixels) between the neighbouring
cutouts. This resulted in a catalog of some 430,000 cutouts for the entire \cfhtls
WIDE region. The size of the individual cutout was determined by
optimising factors such as the typical angular scales of gravitational
lenses, the number of objects seen in a single cutout and the total
number of image cutouts in the survey. If a lens
candidate happens to be too close to the edge of a cutout, then the overlap
between neighbouring cutout allows a volunteer to get a clearer view
of the same candidate in at least one of the cutouts. We note that since
the images are shown randomly, a volunteer may not necessarily come
across the neighbouring cutout unless {\bf they classify} a large
number of images. This is not a problem since our user base is extremely
large and we receive multiple classifications for the same cutout.

@@

All those groups that are found to have no background galaxies within the
cross-sectional area are rejected and the rest are included as potential lenses.
As mentioned earlier, we artificially boost the total number of sources behind
every lens but ensure that the statistical properties such as the profile of the
image separation distribution are not affected (see \Fref{fig:remudist}). We
follow the same procedure and apply thresholds to determine properties
of the lensed galaxies for every lens as are described for galaxy-galaxy lenses
in the previous section. The thresholds are same as those used for
galaxy lenses (see \Tref{tab:thresh}) and are reported within ``()'', if
different for group scales. The simulated lensed images are then added
to the real \cfhtls~images with the BGGs as the center by following
exactly the same procedure as described in the previous section.

@@

\hline
\end{tabular}
\end{center}
{KC}-- Known lens candidates \\
{KL}-- Known lenses \\
{NC}-- New lens candidates  \\
{AC}-- All (known and new) lens candidates  \\
$P_{\rm thresh}$ -- systems with Posterior probability~$P$ above this threshold are selected \\
Note: For KC and KL, Percentages are with respect to the known
population whereas for NC and AC, percentages are with respect to the
total population of lens candidates. \\
\end{table}

@@

We show images of our new sample in \Fref{fig:lc}. The panels are arranged
first in the descending order of their grades, and within each grade, in
ascending order
of RA. As the first lens search was a blind search with no
preselection of candidates from any algorithm, we find various types of
lenses, as expected from such a search. The final sample consists of
both galaxy and group-scale lens candidates. There are detections of
elongated arcs and some interesting point-like quasar lensed images.
Most of them are brighter in the bluer $g$ band, but some candidates brighter
in the redder $i$ band are also found. Since the robotic lens searches
focused on the blue lensed features, they are likely to miss
such interesting lens candidates. We did not find any examples of
exotic lens candidates from the visually inspected $P>0.3$ sample.

@@

We use the publicly available redshifts for the lens galaxy from the
\cfhtls photometric catalogs \citep{Coupon2009}. The total flux of the
lensed image or arc is measured in the $g$ band but the adopted method
is different for different samples. For the simulated sample, we
multiply the magnification of the second brightest image with the source
magnitude. For the \rf sample, the arcs are detected in the scaled
difference image of $g$ and $i$ bands from which the lensing galaxy is
subtracted \citep[for details, see][]{Gavazzi2014}. Here, we use the flux
of the lensed images measured by {\sc sextractor} from the scaled
difference image, that is, $g-\alpha i$ and convert it to the $g$ band
flux using mean colors of the foreground and background population. For
the \af and the \sw sample, we integrate the flux in the image pixels
identified by \af or {\sc sextractor}.

The Einstein radius is also measured differently for
different samples. For the galaxy-scale lenses in the simulated sample,
we use the value of the input parameter of the lens model for the $R_{\rm E}$.
For group-scale lenses, since the lens model is multi-component, we
need to determine the $R_{\rm E}$ from the image positions. We use those
pairs of lensed images that have the smallest and the largest angular
separations. The $R_{\rm E}$ here is then half of the averaged values of
these angular image separations. For the \rf sample, we use the peak
position of the lensed images measured by running {\sc sextractor} on
the scaled difference image. We calculate the image separation from the
lens center as an estimate of the $R_{\rm E}$. For the \af (SARCS)
sample, we use the same definition as above except that the peak
position is identified either by the \af or manually. For the \sw lens
sample, the same definition is used where the peak positions are
identified either with \af or {\sc sextractor}.

@@

Lastly, in \Fref{fig:stackremg}, we show the relative distribution of
number of candidates from each sample as a function of the Einstein
radius and arc magnitude. The light blue color shows the overlap between
the \sw and the \rf samples and the purple color shows the overlap
between the \sw candidates and the \af samples. As noted earlier, the
\rf dominates the small $R_{\rm E} (<2\arcsec)$ range although \sw does
find modest number of candidates in this range. At larger $R_{\rm E}$,
\sw sample begins to dominate and is comparable to the \af sample.  As a
function of the arc magnitudes, all three samples have detections at all
magnitudes and median magnitudes for all samples is around 24.5.
Relatively, the \rf sample spans a narrower range compared to the \sw
and \af sample. However, this can be verified only after understanding
and accounting for the systematic uncertainties in our measurements.

\begin{figure}
\begin{center}
\includegraphics[scale=0.65]{zl_mg_re.pdf}
\caption{ \label{fig:zlmgre}
Comparison of the lens redshift and the arc magnitude with the
Einstein radius for all of the three lens samples, namely, the \rf (green dots),
\sw (known candidates$-$blue circles and new candidates only$-$blue dots)
and \af (red dots). Alls samples have broadly similar properties.}
\end{center}
\end{figure}

@@

\subsection{False negatives: known lenses missed by \sw}
\label{sec:fn}
Like any lens finding method, the \sw system can fail to detect certain
kinds of lenses.  We find that about 40~per~cent of the known sample of lenses
are missed both at \StageOne and II (see \Tref{tab:stats}). In fact, most
of the known lenses from \StageOne are discovered at \StageTwo which
means that the lenses are lost already at \StageOne. Below, we focus on the known
lens sample at \StageOne to understand why some of them are being missed
and possibly find a way to improve the detection rate which can be
adopted in the future \sw lens searches.

@@

For a quantitative comparison of the large and small kicks for the
entire samples of detected and missed known lenses, we show a plot of
histogram on the right of \Fref{fig:detmis}. Qualitatively, there are
four types of volunteers making classifications - those causing positive
large kicks (correct classifications by high power volunteers, those
causing positive small kicks (correct classifications by low power
volunteers), those causing negative small kicks (incorrect
classifications by low power volunteers) and those causing negative
large kicks (incorrect classifications by high power volunteers). The
four histograms in the figure correspond to these four types of
volunteers for each sample (that is, detected or missed). The kick size
is small if $\Delta log{P} = \log{P_{\rm current}} - \log{P_{\rm
previous}} < \Delta\log{P}_{\rm cut}$ (chosen as 1.2) and is large
if greather than $\Delta\log{P}_{\rm cut}$.

@@

We rerun SWAP for \StageOne where we use classifications from volunteers
who produce $\Delta$ log(P)$>1.2$ only. This obviously means reducing
the total Nclass per subject by a large fraction. As a result, we also
need to change the $P_{\rm acc thresh}$ which is chosen to be 0.1 and we
find that about a third of the missed lenses are detected while all the
previously detected lenses remain detected too. The others simply do not
have enough classifications from volunteers producing large positive
kicks. Thus, it may be possible to detect more promising candidates if
we preferentially show the rejected systems to a few high power
volunteers. We understand that changing the rejection and acceptance
thresholds will likely increase the rate of FPs along with improved
completeness. After testing on a subsample of both \StageOne and
\StageOne FPs, we find that a small fraction of FPs are rejected when
only classifications with $\Delta$ log(P)$>1.2$ only are considered.
Therefore, we think it is worth adopting the new strategy in the future
runs of SWAP in future lens searches.

@@

The bottom panels show the missed or rejected lenses. The green (and red)
trajectories show visually easier (and more difficult) to identify
lenses. In spite of some mild qualitative differences, both set of
trajectories have very similar behaviour. The trajectories in panel e
are typical of this sample in terms of Nclass and the dominance of short
negative kicks. The panel f represents a small fraction of this sample
where the kicks are only short and negative. The panel g shows how some
lenses receive a bunch of large positive kicks which are led to
rejection by still mostly negative short kicks. Finally, panel h shows
those cases of lenses which received almost sufficient number of large
positive kicks to be detected but ended up being rejected.

The top panels show the detected lenses. The cases shown with green trajectories
in each panel can be thought of counterparts of the trajectories of the missed
lenses in the corresponding bottom panels except that the type of volunteers are
different. Most detected lenses are similar to the case in panel a which are
detected within a few classifications but coming from large positive kicks.
Panel b represents a few odd cases which are dominated mainly by short positive
kicks. Panel c shows a lens getting more classifications because of the tug
between positive and negative kicks whereas the panel d represents the extreme
cases when the images are on the verge of being rejected but are saved thanks
to a series of large positive kicks. The red trajectories are some more
examples of randomly selected cases which demonstrate how having sufficient
number of large positive kicks allows lenses to be detected in spite of several
short negative kicks.

>>>>

We extracted contiguous cutouts of size 82\arcsec\ (440 pixels), including
overlapping region of 10\arcsec (54 pixels) between the neighbouring
cutouts. This resulted in a catalog of some 430,000 cutouts for the entire \cfhtls
WIDE region. The size of the individual cutout was determined by
optimising factors such as the typical angular scales of gravitational
lenses, the number of objects seen in a single cutout and the total
number of image cutouts in the survey. If a lens
candidate happens to be too close to the edge of a cutout, then the overlap
between neighbouring cutouts allows a volunteer to get a clearer view
of the same candidate in at least one of the cutouts. We note that since
the images are shown randomly, a volunteer may not necessarily come
across the neighbouring cutout unless {\bf they classify} a large
number of images. This is not a problem since our user base is extremely
large and we receive multiple classifications for the same cutout.

@@

All those groups that are found to have no background galaxies within the
cross-sectional area are rejected and the rest are included as potential lenses.
As mentioned earlier, we artificially boost the total number of sources behind
every lens but ensure that the statistical properties such as the profile of the
image separation distribution are not affected (see \Fref{fig:remudist}). We
follow the same procedure and apply thresholds to determine properties
of the lensed galaxies for every lens as described for galaxy-galaxy lenses
in the previous section. The thresholds are same as those used for
galaxy lenses (see \Tref{tab:thresh}) and are reported within ``()'', if
different for group scales. The simulated lensed images are then added
to the real \cfhtls~images with the BGGs as the center by following
exactly the same procedure as described in the previous section.

@@

\hline
\end{tabular}
\end{center}
{KC}-- Known lens candidates \\
{KL}-- Known lenses \\
{NC}-- New lens candidates  \\
{AC}-- All (known and new) lens candidates  \\
$P_{\rm thresh}$ -- systems with Posterior probability~$P$ above this threshold are selected \\
Note: For KC and KL, Percentages are with respect to the known
population (i.e. 438 KC and 59 KL see \Sref{sec:data:kls}) whereas for
NC and AC, percentages are with respect to the total sample of 429 lens
candidates.
\\
\end{table}

@@

We show images of our new sample in \Fref{fig:lc}. The panels are arranged
first in the descending order of their grades, and within each grade, in
ascending order
of RA. As the first lens search was a blind search with no
preselection of candidates from any algorithm, we find various types of
lenses, as expected from such a search. The final sample consists of
both galaxy and group-scale lens candidates. There are detections of
elongated arcs and some interesting point-like quasar lensed images.
Most of them are brighter in the bluer $g$ band, but some candidates brighter
in the redder $i$ band are also found. Since the robotic lens searches
focused on the blue lensed features, they are likely to miss
such interesting lens candidates. We did not find any examples of
exotic lens candidates from the visually inspected $P>0.3$ sample.
There
may be some more interesting candidates that were missed either at
\StageOne or \StageTwo but have been identified by volunteers in \Talk.
This resource is yet to be mined and is left for future work.

@@

We use the publicly available redshifts for the lens galaxy from the
\cfhtls photometric catalogs \citep{Coupon2009}. The total flux of the
lensed image or arc is measured in the $g$ band but the adopted method
is different for different samples. For the simulated sample, we
multiply the magnification of the second brightest image with the source
magnitude. For the \rf sample, the arcs are detected in the scaled
difference image of $g$ and $i$ bands from which the lensing galaxy is
subtracted \citep[for details, see][]{Gavazzi2014}. Here, we use the flux
of the lensed images measured by {\sc SExtractor} from the scaled
difference image, that is, $g-\alpha i$ and convert it to the $g$ band
flux using mean colors of the foreground and background population. For
the \af and the \sw sample, we integrate the flux in the image pixels
identified by \af or {\sc SExtractor}.

The Einstein radius is also measured differently for
different samples. For the galaxy-scale lenses in the simulated sample,
we use the value of the input parameter of the lens model for the $R_{\rm E}$.
For group-scale lenses, since the lens model is multi-component, we
need to determine the $R_{\rm E}$ from the image positions. We use those
pairs of lensed images that have the smallest and the largest angular
separations. The $R_{\rm E}$ here is then half of the averaged values of
these angular image separations. For the \rf sample, we use the peak
position of the lensed images measured by running {\sc SExtractor} on
the scaled difference image. We calculate the image separation from the
lens center as an estimate of the $R_{\rm E}$. For the \af (SARCS)
sample, we use the same definition as above except that the peak
position is identified either by the \af or manually. For the \sw lens
sample, the same definition is used where the peak positions are
identified either with \af or {\sc SExtractor}.

@@

Lastly, in \Fref{fig:stackremg}, we show the relative distribution of
number of candidates from each sample as a function of the Einstein
radius and arc magnitude. The light blue color shows the overlap between
the \sw and the \rf samples and the purple color shows the overlap
between the \sw candidates and the \af samples. As noted earlier, the
\rf dominates the small $R_{\rm E} (<2\arcsec)$ range although \sw does
find modest number of candidates in this range. At larger $R_{\rm E}$,
\sw sample begins to dominate and is comparable to the \af sample.  As a
function of the arc magnitudes, all three samples have detections at all
magnitudes and median magnitudes for all samples is around $g\sim$24.5.
Relatively, the \rf sample spans a narrower range compared to the \sw
and \af sample. However, this can be verified only after understanding
and accounting for the systematic uncertainties in our measurements.

\begin{figure}
\begin{center}
\includegraphics[scale=0.65]{zl_mg_re.pdf}
\caption{ \label{fig:zlmgre}
Comparison of the lens redshift and the arc magnitude with the
Einstein radius for all of the three lens samples, namely, the \rf (green dots),
\sw (known candidates$-$blue circles and new candidates only$-$blue dots)
and \af (red dots). All samples have broadly similar properties.}
\end{center}
\end{figure}

@@

\subsection{False negatives: known lenses missed by \sw}
\label{sec:fn}
Like any lens finding method, the \sw system can fail to detect certain
kinds of lenses.  We find that about 40~per~cent of the known sample of lenses
are missed both at \StageOne and II (see \Tref{tab:stats}). In fact, all
of the known lenses from \StageOne are discovered at \StageTwo which
means that the lenses are lost already at \StageOne. Below, we focus on the known
lens sample at \StageOne to understand why some of them are being missed
and possibly find a way to improve the detection rate which can be
adopted in the future \sw lens searches.

@@

For a quantitative comparison of the large and small kicks for the
entire samples of detected and missed known lenses, we show a plot of
histogram on the right of \Fref{fig:detmis}. Qualitatively, there are
four types of volunteers making classifications - those causing positive
large kicks (correct classifications by high power volunteers), those
causing positive small kicks (correct classifications by low power
volunteers), those causing negative small kicks (incorrect
classifications by low power volunteers) and those causing negative
large kicks (incorrect classifications by high power volunteers). The
four histograms in the figure correspond to these four types of
volunteers for each sample (that is, detected or missed). The kick size
is small if $\Delta log{P} = \log{P_{\rm current}} - \log{P_{\rm
previous}} < \Delta\log{P}_{\rm cut}$ (chosen as 1.2) and is large
if greather than $\Delta\log{P}_{\rm cut}$.

@@

We rerun SWAP for \StageOne where we use classifications from volunteers
who produce $\Delta$ log(P)$>1.2$ only. This obviously means reducing
the total Nclass per subject by a large fraction. As a result, we also
need to change the $P_{\rm acc thresh}$ which is chosen to be 0.1 and we
find that about a third of the missed lenses are detected while all the
previously detected lenses remain detected too. The others simply do not
have enough classifications from volunteers producing large positive
kicks. Thus, it may be possible to detect more promising candidates if
we preferentially show the rejected systems to a few high power
volunteers. We understand that changing the rejection and acceptance
thresholds will likely increase the rate of FPs along with improved
completeness. After testing on a subsample of both \StageOne and
\StageTwo FPs, we find that a small fraction of FPs are rejected when
only classifications with $\Delta$ log(P)$>1.2$ only are considered.
Therefore, we think it is worth adopting the new strategy in the future
runs of SWAP in future lens searches.

@@

The {\bf bottom-left panels of \Fref{fig:detmis} show the} missed or rejected lenses. The green (and red)
trajectories show visually easier (and more difficult) to identify
lenses. In spite of some mild qualitative differences, both set of
trajectories have very similar behaviour. The trajectories in panel e
are typical of this sample in terms of Nclass and the dominance of short
negative kicks. Panel f represents a small fraction of this sample
where the kicks are only short and negative. The panel g shows how some
lenses receive a bunch of large positive kicks which are led to
rejection by still mostly negative short kicks. Finally, panel h shows
those cases of lenses which received almost sufficient number of large
positive kicks to be detected but ended up being rejected.

The {\bf top-left panels of \Fref{fig:detmis}} show the detected lenses. The cases shown with green trajectories
in each panel can be thought of counterparts of the trajectories of the
missed lenses in the corresponding bottom panels except that the type of
volunteers are different. Most detected lenses are similar to the case
in panel a that are detected within a few classifications coming from
large positive kicks.  Panel b represents a few odd cases which are
dominated mainly by short positive kicks. Panel c shows a lens getting
more classifications but do not reach the detection or rejection
thresholds because of the tug between positive and negative kicks mostly
from experienced volunteers.  Panel d represents the extreme cases when
the images are on the verge of being rejected but are saved thanks to a
series of large positive kicks.  The red trajectories are some more
examples of randomly selected cases which demonstrate how having
sufficient number of large positive kicks allows lenses to be detected
in spite of several short negative kicks.
